How to improve PCA based methods of genome scan using ecological data: detecting selection using RDA.

### 2017-04-19 article_genome_scan.tex
update

From the analysis of \textit{Popular trichocarpa}, we picked up some genes that can be interpreted regarding to the environmental variable. For instance, among the top SNPs, some of them are located within or nearby, POPTR\_0015s00640 implicated in cellular response to cold, POPTR\_0015s00440 involved in the regulation of circadian clock in plants. One outlier, POPTR\_0017s06840 that is not detected by pcadapt is involved in amylase catabolism, an activity that is suspected to depend of climatic conditions. Interestingly, Table S2 shows that a substantial amount of RDA genome scan outliers (35) are not detected by any other genome scan method including pcadapt. For instance, four SNPs are assigned to POPTR\_0007s04340, a dehydration-responsive protein-related which is similar to similar to early-responsive to dehydration stress ERD3 protein. However, its function remains mainly unknown. These results support the simulation analysis conclusions that RDA based genome scan is able to pick up more genes in relation to some environmental gradient than pcadapt. It is however noticeable that a large proportion of outliers are shared between PCA and RDA based methods which highlight the fact that both approaches are very similar. The lack of interpretability of PCA method is compensated by the fact that it is a blind method in regard to environmental data so it can pick up genes that we will miss using RDA because some crucial environmental data are lacking.
We can notice that there is a strong climate environmental gradient structuring the adaptive genetic space represented by RDA1. This is not surprising that so many outliers are shared between pcadapt and our RDA based genome scan method. Indeed, all climate variable are autocorrelated to the latitudinal gradient where this species has been sampled. RDA2 is not very informative as it does not clearly seperate a peculiar set of orthogonal environmental variable except PAS which stands for precipitation as snow. A substantial number of SNPs are clearly correlated to RDA1 gradient, most of them being detected by both methods. Furthermore, when looking at the proportion of variance explained by the first three axis, Fig. \ref{varexplainedPopulus} clearly confirms our simulation conclusion by showing that the adaptive genetic space is better aligned to the environmental variable then the whole genetic space. RDA1 can therefore serve as a basis to design a composite index of climate adaptation as it was proposed by \citet{Steane2014a}. This index can serve to predict (i) provenances that would perform well in the common gardens and (ii) patterns of adaptation to local climate across the geographic range.
In conclusion, although there is a large covariance between climatic variable, this analysis proves that RDA can be used to identify environmental gradients that are responsible for the adaptive differentiation across the geographic distribution of \textit{Populus trichocarpa}. This was intuitively expected but the approach is now validated on simulated datasets.
Finally, this paper shows that both methods are complementary and should be used simultaneously on the same dataset. RDA will be more easily interpretable in terms of mechanism and should pick up more genes when some important selective gradient are identified but PCA might be able to pick up outliers where no environmental data are available. However this latter method will miss outliers when selective gradient are poorly correlated to the population structure (i.e. geographical distance between populations and individuals). It is therefore crucial when performing landscape genomics to clearly characterize the environmental condition by measuring important variable that are suspected to put selective pressures on the species of interest. This can be done through RDA as it has been validated in our simulations and in the real dataset on \textit{Populus trichocarpa}.

>>>>

From the analysis of \textit{Popular trichocarpa}, we picked up some genes that can be interpreted regarding to the environmental variable. For instance, among the top SNPs, some of them are located within or nearby, POPTR\_0015s00640 implicated in cellular response to cold, POPTR\_0015s00440 involved in the regulation of circadian clock in plants. One outlier, POPTR\_0017s06840 that is not detected by pcadapt is involved in amylase catabolism, an activity that is suspected to depend of climatic conditions. Interestingly, Table S2 shows that a substantial amount of RDA genome scan outliers (35) are not detected by any other genome scan method including pcadapt. For instance, four SNPs are assigned to POPTR\_0007s04340, a dehydration-responsive protein-related which is similar to similar to early-responsive to dehydration stress ERD3 protein. However, its function remains mainly unknown. These results support the simulation analysis conclusions that RDA based genome scan is able to pick up more genes in relation to some environmental gradient than pcadapt. It is however noticeable that a large proportion of outliers are shared between PCA and RDA based methods which highlight the fact that both approaches are very similar. The lack of interpretability of PCA method is compensated by the fact that it is a blind method in regard to environmental data so it can pick up genes that we will miss using RDA because some crucial environmental data are lacking.

We can notice that there is a strong climate environmental gradient structuring the adaptive genetic space represented by RDA1 (Fig. \ref{fig:poplar_rda}). This is not surprising that so many outliers are shared between pcadapt and our RDA based genome scan method. Indeed, all climate variable are autocorrelated to the latitudinal gradient where this species has been sampled. RDA2 is not very informative as it does not clearly seperate a peculiar set of orthogonal environmental variable except PAS which stands for precipitation as snow. A substantial number of SNPs are clearly correlated to RDA1 gradient, most of them being detected by both methods. Furthermore, when looking at the proportion of variance explained by the first three axis, Fig. \ref{fig:varexplainedPopulus} clearly confirms our simulation conclusion by showing that the adaptive genetic space is better aligned to the environmental variable then the whole genetic space. RDA1 can therefore serve as a basis to design a composite index of climate adaptation as it was proposed by \citet{Steane2014a}. This index can serve to predict (i) provenances that would perform well in the common gardens and (ii) patterns of adaptation to local climate across the geographic range.

In conclusion, although there is a large covariance between climatic variable, this analysis proves that RDA can be used to identify environmental gradients that are responsible for the adaptive differentiation across the geographic distribution of \textit{Populus trichocarpa}. This was intuitively expected but the approach is now validated on simulated datasets.

Finally, this paper shows that both methods are complementary and should be used simultaneously on the same dataset. RDA will be more easily interpretable in terms of mechanism and should pick up more genes when some important selective gradient are identified but PCA might be able to pick up outliers where no environmental data are available. However this latter method will miss outliers when selective gradient are poorly correlated to the population structure (i.e. geographical distance between populations and individuals). It is therefore crucial when performing landscape genomics to clearly characterize the environmental condition by measuring important variable that are suspected to put selective pressures on the species of interest. This can be done through RDA as it has been validated in our simulations and in the real dataset on \textit{Populus trichocarpa}.
